---
title: "Perform ClusterDE on a spatial dataset (two domains)"
author:
  - name: Siqi Chen
    affiliation:
    - Computer Science, Central South University
    email: siqichen4477@gmail.com
  - name: Dongyuan Song
    affiliation:
    - Department of Genetics & Genome Sciences, UConn Health 
    - Bioinformatics IDP, University of California, Los Angeles
    email: dongyuansong@ucla.edu
output:
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
package: "`r pkg_ver('ClusterDE')`"
vignette: >
  %\VignetteIndexEntry{ClusterDE-spatial-twodomains_vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
tools::R_user_dir("ClusterDE", which="cache")
```

# Download data

We selected adjacent layer 6 and WM of 151673 slice from the LIBD Human Dorsolateral Prefrontal Cortex (DLPFC) dataset, which is downloaded in the spatialLIBD R package. We removed the genes expressed in fewer than 20% spots to save computational time here.
```{r}
data(dlpfc_twodomain, package = "ClusterDE")
```

Visualize the real data with two domains (L6 and WM).
```{r}
# Visualize the real spatial domains
domains <- data.frame(Xaxis = dlpfc_twodomain$spatial1, Yaxis = dlpfc_twodomain$spatial2, Domains = dlpfc_twodomain$cell_type)
ggplot2::ggplot(domains, ggplot2::aes(x = Xaxis, y = Yaxis, col = Domains)) +
  ggplot2::geom_point(size = 1.0) +
  ggplot2::coord_equal() +
  ggplot2::ggtitle("Manual annotation \n (The layer6 and WM in the slice 151673)") +
  ggplot2::theme(
    plot.title = ggplot2::element_text(size = 10, hjust = 0.5),
    panel.grid = ggplot2::element_blank(),
    panel.background = ggplot2::element_rect(fill = "gray90"),
    panel.border = ggplot2::element_rect(color = "black", fill = NA, linewidth = 0.6),
    axis.title.x = ggplot2::element_blank(),
    axis.title.y = ggplot2::element_blank(),
    axis.ticks.x = ggplot2::element_blank(),
    axis.ticks.y = ggplot2::element_blank(),
    axis.text.x = ggplot2::element_blank(),
    axis.text.y = ggplot2::element_blank()
  ) +
  ggplot2::scale_color_manual(values = c("#5791cc", "#403f92"))
```

# Run the BayesSpace + Seurat pipeline

First, we employed the BayesSpace for spatial clustering. Please note that ClusterDE is designed for 1 vs 1 comparison; therefore, we obtain two spatial clusters for illustration purpose.
```{r}
# Construct the input of BayesSpace based on real dataset
# The input of BayesSpace is sce object
dlpfc_twodomain_sce <- SingleCellExperiment::SingleCellExperiment(
  list(counts = Seurat::GetAssayData(dlpfc_twodomain, layer = "counts"))
)
# Add colData information of SingleCellExperiment
dlpfc_twodomain_sce$row <- dlpfc_twodomain@meta.data$row
dlpfc_twodomain_sce$col <- dlpfc_twodomain@meta.data$col
dlpfc_twodomain_sce$spatial1 <- dlpfc_twodomain@meta.data$spatial1
dlpfc_twodomain_sce$spatial2 <- dlpfc_twodomain@meta.data$spatial2

# Log-normalize the count data
RNGkind("L'Ecuyer-CMRG")
seed <- 123
set.seed(seed)
dlpfc_twodomain_sce <- BayesSpace::spatialPreprocess(
  dlpfc_twodomain_sce,
  platform = "Visium",
  n.PCs = 7,
  log.normalize = T
)
# Clustering with BayesSpace
dlpfc_twodomain_sce <- BayesSpace::spatialCluster(
  dlpfc_twodomain_sce,
  q = 2,
  platform = "Visium",
  d = 7,
  init.method = "mclust",
  model = "t",
  gamma = 2,
  nrep = 1000,
  burn.in = 100,
  save.chain = T
)
dlpfc_twodomain <- Seurat::as.Seurat(dlpfc_twodomain_sce)
Seurat::Idents(dlpfc_twodomain) <- "spatial.cluster"
```

Visualize the spatial clustering results based on the real data.
```{r}
# Visualize the spatial cluster
clusters <- data.frame(
  Xaxis = dlpfc_twodomain$spatial1,
  Yaxis = dlpfc_twodomain$spatial2,
  Clusters = as.character(dlpfc_twodomain$spatial.cluster)
)
ggplot2::ggplot(clusters, ggplot2::aes(x = Xaxis, y = Yaxis, col = Clusters)) +
  ggplot2::geom_point(size = 1.0) +
  ggplot2::coord_equal() +
  ggplot2::ggtitle("Real data \n (spatial clusters detected by BayesSpace)") +
  ggplot2::theme(
    plot.title = ggplot2::element_text(size = 10, hjust = 0.5),
    panel.grid = ggplot2::element_blank(),
    panel.background = ggplot2::element_rect(fill = "gray90"),
    panel.border = ggplot2::element_rect(color = "black", fill = NA, size = 0.6),
    axis.title.x = ggplot2::element_blank(),
    axis.title.y = ggplot2::element_blank(),
    axis.ticks.x = ggplot2::element_blank(),
    axis.ticks.y = ggplot2::element_blank(),
    axis.text.x = ggplot2::element_blank(),
    axis.text.y = ggplot2::element_blank()
  ) +
  ggplot2::scale_color_manual(values = c("#e87d72", "#54bcc2"))
```

Then, we used the common DE method (Wilcoxon Rank Sum Test) to identify domain marker genes between the two spatial clusters.
```{r}
original_deg <- Seurat::FindMarkers(
  object = dlpfc_twodomain,
  ident.1 = 2,
  ident.2 = 1,
  test.use = "wilcox",
  logfc.threshold = 0,
  min.pct = 0,
  min.cells.feature = 1,
  min.cells.group = 1
)
original_pval <- original_deg$p_val_adj
names(original_pval) <- rownames(original_deg)
print(paste0("Seurat found ", sum(original_deg$p_val_adj < 0.05), " DEG"))
```

# Find DEGs using ClusterDE

We use spatial coordinates as covariates to generate synthetic null data, then apply the same BayesSpace clustering method as target data.
```{r}
set.seed(seed)
null_data <- ClusterDE::constructNull(
  dlpfc_twodomain,
  data_type = "spatial",
  formula = "s(spatial1, spatial2, bs = 'gp', k = 4)",
  other_covariates = c("spatial1", "spatial2")
)
null_obj_sce <- SingleCellExperiment::SingleCellExperiment(list(counts = null_data))
null_obj_sce$row <- dlpfc_twodomain@meta.data$row
null_obj_sce$col <- dlpfc_twodomain@meta.data$col
null_obj_sce$spatial1 <- dlpfc_twodomain@meta.data$spatial1
null_obj_sce$spatial2 <- dlpfc_twodomain@meta.data$spatial2
null_obj_sce <- BayesSpace::spatialPreprocess(
  null_obj_sce,
  platform = "Visium",
  n.PCs = 7,
  log.normalize = T
)
null_obj_sce <- BayesSpace::spatialCluster(
  null_obj_sce,
  q = 2,
  platform = "Visium",
  d = 7,
  init.method = "mclust",
  model = "t",
  gamma = 2,
  nrep = 1000,
  burn.in = 100,
  save.chain = T
)
null_obj <- Seurat::as.Seurat(null_obj_sce)
Seurat::Idents(null_obj) <- "spatial.cluster"
```

Compare gene p-values from target and synthetic data to calculate ClusterDE FDR controlled DEG.
```{r}
null_deg <- Seurat::FindMarkers(
  object = null_obj,
  ident.1 = 1,
  ident.2 = 2,
  test.use = "wilcox",
  logfc.threshold = 0,
  min.pct = 0,
  min.cells.feature = 1,
  min.cells.group = 1
)
null_pval <- null_deg$p_val_adj
names(null_pval) <- rownames(null_deg)

deg <- ClusterDE::callDE(original_pval, null_pval)
print(paste0("ClusterDE found ", sum(deg$record >= 0.5), " DEG"))
```

# Session information

```{r}
sessionInfo()
```
