---
title: "Perform ClusterDE on a monocyte scRNA dataset"
author:
  - name: Dongyuan Song
    affiliation:
    - Department of Genetics & Genome Sciences, UConn Health 
    - Bioinformatics IDP, University of California, Los Angeles
    email: dongyuansong@ucla.edu
output:
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
package: "`r pkg_ver('ClusterDE')`"
vignette: >
  %\VignetteIndexEntry{ClusterDE-monocyte-scrna_vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
tools::R_user_dir("ClusterDE", which = "cache")
```

# Download data

The PBMC datasets are originally from `SeuratData`. We use one of them (10x Chromium (v3) from PBMC1 replicate). We filtered out some lowly epxressed genes to save computational time here.
```{r}
data(pbmc, package = "ClusterDE")
```

# Run the regular Seurat pipeline

We perform the default Seurat clustering. Note that in real data analysis, the cell type label is usually unknown.
```{r fig.width=12, fig.height=6}
RNGkind("L'Ecuyer-CMRG")
seed <- 123
set.seed(seed)
pbmc <- Seurat::UpdateSeuratObject(pbmc)
pbmc <- Seurat::NormalizeData(pbmc)
pbmc <- Seurat::FindVariableFeatures(pbmc)
pbmc <- Seurat::ScaleData(pbmc)
pbmc <- Seurat::RunPCA(pbmc)
pbmc <- Seurat::FindNeighbors(pbmc)
pbmc <- Seurat::FindClusters(pbmc, resolution = 0.3)
pbmc <- Seurat::RunUMAP(pbmc, dims = 1:10)
p1 <- Seurat::DimPlot(pbmc, reduction = "umap", label = T) +
  ggplot2::ggtitle("Clustering result") +
  Seurat::NoLegend()
p2 <- Seurat::DimPlot(pbmc, reduction = "umap", group.by = "CellType", label = T) +
  Seurat::NoLegend()
p1 + p2
```

In this vignette, we are interested in cluster 3 vs 8, which approximately represent CD14+/CD16+ monocytes. Please note that ClusterDE is designed for 1 vs 1 comparison. Therefore, users may (1) choose the two interested clusters manually based on their knowledge or (2) use the two locally closest clusters from computation (e.g., `BuildClusterTree` in `Seurat`).
```{r}
library(Seurat)
pbmc <- BuildClusterTree(pbmc)
PlotClusterTree(pbmc)
```

We subset the cluster 3 and 8 (`pbmc_sub`) for further analysis.
```{r}
pbmc_sub <- subset(x = pbmc, idents = c(3, 8))
# Remove genes with zero variance
non_zero_genes <- apply(Seurat::GetAssayData(pbmc_sub, layer = "counts"), 1, var) != 0
pbmc_sub <- pbmc_sub[non_zero_genes,]
```

# Find DEGs using ClusterDE

First, we generate a synthetic null data from the target pbmc subset data.
```{r}
null_data <- ClusterDE::constructNull(pbmc_sub)
```

Next we perform the same preprocess pipeline for the null data as the target data, which apply the standard Seurat pipeline to log normalize counts and find cell clusters. We set the resolution to be 0.3 to find 2 clusters for the null data.
```{r fig.width=12, fig.height=6}
set.seed(seed)
null_obj <- Seurat::CreateSeuratObject(null_data)
null_obj <- Seurat::NormalizeData(null_obj)
null_obj <- Seurat::FindVariableFeatures(null_obj)
null_obj <- Seurat::ScaleData(null_obj)
null_obj <- Seurat::RunPCA(null_obj)
null_obj <- Seurat::FindNeighbors(null_obj)
null_obj <- Seurat::FindClusters(null_obj, resolution = 0.3)
```

Then we perform DE analysis on both target data and synthetic null data using Seurat methods.
```{r}
original_deg <- Seurat::FindMarkers(
  pbmc_sub,
  ident.1 = 3,
  ident.2 = 8,
  min.pct = 0,
  logfc.threshold = 0
)
original_pval <- original_deg$p_val_adj
names(original_pval) <- rownames(original_deg)
null_deg <- Seurat::FindMarkers(
  null_obj,
  ident.1 = 0,
  ident.2 = 1,
  min.pct = 0,
  logfc.threshold = 0
)
null_pval <- null_deg$p_val_adj
names(null_pval) <- rownames(null_deg)
```

Finally, compare original p-values and null p-values using ClusterDE FDR control to find DE.
```{r}
deg <- ClusterDE::callDE(original_pval, null_pval)
print(paste0("Seurat pipeline found ", sum(original_deg$p_val_adj < 0.05), " DEG"))
print(paste0("ClusterDE found ", sum(deg$record >= 0.5), " DEG"))
```

To compare the result from the naive Seurat pipeline and ClusterDE, we visualize the top 6 DE genes from Seurat. Genes *LST1* and *RPS19* are both highly expressed in two clusters. In addition, *RPS19* is reported as a stable housekeeping genes in several studies. Note that it does not mean the expression levels of *LST1* and *RPS19* are the same between the two cell types. It means that they are not good cell type markers. Philosophically speaking, it means that **conditional on the two clusters are obtained by clustering algorithm, LST1 and RPS19 are less likely to be the cell type markers between the two cell types**.
```{r, fig.width=10, fig.height=6}
Seurat::FeaturePlot(pbmc_sub, features = rownames(original_deg)[1:6], ncol = 3)
```

In contrast, the genes from ClusterDE do not have *LST1* and *RPS19* anymore.
```{r, fig.width=10, fig.height=6}
Seurat::FeaturePlot(pbmc_sub, features = deg$gene[1:6], ncol = 3)
```

# Session information
```{r}
sessionInfo()
```
