---
title: "Perform ClusterDE on a pure cell-line dataset"
author:
  - name: Dongyuan Song
    affiliation:
    - Department of Genetics & Genome Sciences, UConn Health 
    - Bioinformatics IDP, University of California, Los Angeles
    email: dongyuansong@ucla.edu
output:
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
package: "`r pkg_ver('ClusterDE')`"
vignette: >
  %\VignetteIndexEntry{ClusterDE-pure-cellline_vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
tools::R_user_dir("ClusterDE", which = "cache")
```

# Download data

We download the cell line data set H2228. The original data is from [Tian et al., Nature Methods 2019](https://github.com/LuyiTian/sc_mixology) as the gold standard for benchmarking the accuracy of clustering. Since the data is from the pure cell line, it should not have cell types, and, of course, between cell type DE genes. **Note**: it does not mean that there are no variations within one cell line; of course there are, e.g., cell cycle, total UMI, etc. However, these variations do not represent **discrete cell groups**, and essentially it means you should not use your obtained clusters to explain the variation.
```{r}
data(cellline, package = "ClusterDE")
```

# Run the regular Seurat pipeline

We perform the default Seurat clustering. Please note that ClusterDE is designed for 1 vs 1 comparison; therefore, we set the resolution as 0.2 here to obtain two clusters for illustration purpose.
```{r}
RNGkind("L'Ecuyer-CMRG")
seed <- 123
set.seed(seed)
cellline <- Seurat::UpdateSeuratObject(cellline)
cellline <- Seurat::NormalizeData(cellline)
cellline <- Seurat::FindVariableFeatures(cellline)
cellline <- Seurat::ScaleData(cellline)
cellline <- Seurat::RunPCA(cellline)
cellline <- Seurat::FindNeighbors(cellline)
cellline <- Seurat::FindClusters(cellline, resolution = 0.2)
cellline <- Seurat::RunUMAP(cellline, dims = 1:10)
Seurat::DimPlot(cellline, reduction = "umap") + ggplot2::ggtitle("Clustering result")
```

From the UMAP, the two clusters seem to be dubious. Although we do not expect the existence of cell types, when we perform Seurat DE test between the two clusters, and we get DE genes with FDR < 0.05. It means that the double-dipping introduces a huge number of false discoveries.

```{r}
original_deg <- Seurat::FindMarkers(
  cellline,
  ident.1 = 0,
  ident.2 = 1,
  min.pct = 0,
  logfc.threshold = 0
)
original_deg <- original_deg[original_deg$avg_log2FC > 0,]
original_pval <- original_deg$p_val_adj
names(original_pval) <- rownames(original_deg)
print(paste0("Seurat found ", sum(original_deg$p_val_adj < 0.05), " DEG"))
```

# Find DEGs using ClusterDE

We generate the synthetic null replicate from the target data using `ClusterDE::constructNULL` and perform the same preprocessing and clustering pipeline as the target cell line data.
```{r}
set.seed(seed)
null_data <- ClusterDE::constructNull(cellline, data_type = "cellline", seed = seed)
null_obj <- Seurat::CreateSeuratObject(null_data)
null_obj <- Seurat::NormalizeData(null_obj)
null_obj <- Seurat::FindVariableFeatures(null_obj)
null_obj <- Seurat::ScaleData(null_obj)
null_obj <- Seurat::RunPCA(null_obj)
null_obj <- Seurat::FindNeighbors(null_obj)
null_obj <- Seurat::FindClusters(null_obj, resolution = 0.3)
```

We then perform Wilcoxon DE test on the null data and compare it with the p-values of target genes. We can observe that after ClusterDE FDR control, no genes pass the test which means we don't detect any DEG in this cell line data.
```{r}
null_deg <- Seurat::FindMarkers(
  null_obj,
  ident.1 = 0,
  ident.2 = 1,
  min.pct = 0,
  logfc.threshold = 0
)
null_deg <- null_deg[null_deg$avg_log2FC > 0,]
null_pval <- null_deg$p_val_adj
names(null_pval) <- rownames(null_deg)
deg <- ClusterDE::callDE(original_pval, null_pval)
print(paste0("ClusterDE found ", sum(deg$record >= 0.5), " DEG"))
```

# Session information
```{r}
sessionInfo()
```
