---
title: "Perform ClusterDE on a single-cell multiome dataset"
author:
  - name: Yihui Cen
    email: ycen@fredhutch.org
output:
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
package: "`r pkg_ver('ClusterDE')`"
vignette: >
  %\VignetteIndexEntry{ClusterDE-scmultiome-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
tools::R_user_dir("ClusterDE", which = "cache")
```

# Download data

Here, we load single-cell multiome PBMC dataset done by 10X Genomics, originally from SeuratData. Specifically, we extract the CD4 Naive and CD4 TCM cell types and randomly sampled 500 cells for each cell type. To further save computational time, features with relatively low expression are filtered out.  All the cells are paired in both the RNA and ATAC modalities.
```{r, results = "hide", message = F}
data(pbmc_rna, package = "ClusterDE")
table(Seurat::Idents(pbmc_rna))
data(pbmc_atac, package = "ClusterDE")
table(Seurat::Idents(pbmc_atac))
```

# Generate synthetic null data for the RNA data

We first check on the RNA data. Generate 20 null replicates using ClusterDE stable to eliminate randomness of null generation.
```{r}
RNGkind("L'Ecuyer-CMRG")
seed <- 123
set.seed(seed)
null_data <- ClusterDE::constructNull(
  obj = pbmc_rna,
  n_rep = 20,
  seed = seed
)
```

Next, we run the seurat pipeline on the null data
```{r, results = "hide", message = F}
set.seed(seed)

null_seurat <- lapply(seq_along(null_data), function(i) {
  obj <- Seurat::CreateSeuratObject(
    counts = null_data[[i]],
    assay = "RNA",
    project = paste0("null_", i)
  )

  obj <- Seurat::NormalizeData(
    obj,
    normalization.method = "LogNormalize",
    scale.factor = 10000,
    verbose = FALSE
  )

  obj <- Seurat::FindVariableFeatures(
    obj,
    selection.method = "vst",
    nfeatures = 3000,
    verbose = FALSE
  )

  obj <- Seurat::ScaleData(
    obj,
    features = Seurat::VariableFeatures(obj),
    verbose = FALSE
  )

  obj <- Seurat::RunPCA(
    obj,
    features = Seurat::VariableFeatures(obj),
    npcs = 50,
    reduction.name = "pca",
    reduction.key = "PC_",
    verbose = FALSE
  )

  obj <- Seurat::FindNeighbors(
    obj,
    reduction = "pca",
    dims = 1:30,
    verbose = FALSE
  )
  
  obj <- Seurat::FindClusters(
    obj,
    resolution = 0.4
  )

  obj
})

names(null_seurat) <- paste0("null_", seq_along(null_seurat))
```

# Find DEGs on RNA data

To find DEGs with ClusterDE, we first find DEGs with the original data and obtain the target scores.
```{r}
original_levels <- levels(
  droplevels(Seurat::Idents(pbmc_rna))
)

original_markers <- Seurat::FindMarkers(
  object = pbmc_rna,
  assay = "RNA",
  ident.1 = original_levels[1],
  ident.2 = original_levels[2],
  test.use = "wilcox",
  min.pct = 0,
  logfc.threshold = 0,
  only.pos = FALSE
)

print(paste0("Wilcox found ", sum(p.adjust(original_markers$p_val, method = "BH") < 0.05), " DE"))
```

Then, we find DEGs for the synthetic null data
```{r}
null_markers_list <- lapply(
  null_seurat,
  function(null_obj) {
    null_levels <- levels(
      droplevels(Seurat::Idents(null_obj))
    )

    null_markers <- Seurat::FindMarkers(
      object = null_obj,
      assay = "RNA",
      ident.1 = null_levels[1],
      ident.2 = null_levels[2],
      test.use = "wilcox",
      min.pct = 0,
      logfc.threshold = 0,
      only.pos = FALSE
    )

    null_markers$gene <- rownames(null_markers)

    null_markers
  }
)

names(null_markers_list) <- names(null_seurat)

nullScores_raw <- lapply(
  null_markers_list,
  function(null_markers) {

    scores <- null_markers$p_val

    names(scores) <- rownames(null_markers)

    scores[
      is.finite(scores)
    ]
  }
)
```

After that, we use ClusterDE to call DEG results on the RNA data.
```{r}
original_markers$gene <- rownames(original_markers)
targetScores <- original_markers$p_val
names(targetScores) <- rownames(original_markers)

de <- ClusterDE::callDE(
  targetScores = targetScores,
  nullScores = nullScores_raw,
  FDR = 0.05,
  correct = TRUE,
  nlogTrans = TRUE,
  threshold = "DS"
)

print(paste0("ClusterDE found ", sum(de$record >= 0.5), " DE"))
```

# Generate synthetic null data for the ATAC data

Then, since the RNA data shows clear two clusters, we check on the ATAC data for differential accessibility analysis. Here, we generate 20 null replicates using ClusterDE stable to eliminate randomness of null generation.
```{r}
null_data_atac <- ClusterDE::constructNull(
  obj = pbmc_atac,
  data_type = "scATAC",
  n_rep = 20,
  seed = seed
)
```

Next, we run the seurat pipeline on the null data
```{r, results = "hide", message = F}
set.seed(seed)

null_signac <- lapply(seq_along(null_data_atac), function(i) {
  atac_assay <- Signac::CreateChromatinAssay(
    counts = null_data_atac[[i]],
    sep = c(":", "-"),
    min.cells = 0,
    min.features = 0
  )

  obj <- Seurat::CreateSeuratObject(
    counts = atac_assay,
    assay = "ATAC",
    project = paste0("null_atac_", i)
  )

  obj <- Signac::RunTFIDF(
    obj
  )

  obj <- Signac::FindTopFeatures(
    obj,
    min.cutoff = "q0"
  )

  obj <- Signac::RunSVD(
    obj,
    assay = "ATAC",
    n = 50,
    reduction.name = "lsi",
    reduction.key = "LSI_"
  )

  obj <- Seurat::FindNeighbors(
    obj,
    reduction = "lsi",
    dims = 2:30,
    verbose = FALSE
  )

  obj <- Seurat::FindClusters(
    obj,
    resolution = 0.7
  )

  obj
})

names(null_signac) <- paste0(
  "null_atac_",
  seq_along(null_signac)
)
```

# Find DE peaks on ATAC data

To find DE peaks with ClusterDE, we first find DE peaks with the original data and obtain the target scores.
```{r}
original_levels_atac <- levels(
  droplevels(Seurat::Idents(pbmc_atac))
)

original_markers_atac <- Seurat::FindMarkers(
  object = pbmc_atac,
  assay = "ATAC",
  ident.1 = original_levels_atac[1],
  ident.2 = original_levels_atac[2],
  test.use = "wilcox",
  min.pct = 0,
  logfc.threshold = 0,
  only.pos = FALSE
)

print(
  paste0(
    "Wilcox found ",
    sum(
      p.adjust(
        original_markers_atac$p_val,
        method = "BH"
      ) < 0.05
    ),
    " DA peaks"
  )
)
```

Then, we find DE peaks for the synthetic null data
```{r}
null_markers_atac_list <- lapply(
  null_signac,
  function(null_obj) {
    null_levels <- levels(
      droplevels(Seurat::Idents(null_obj))
    )

    null_markers <- Seurat::FindMarkers(
      object = null_obj,
      assay = "ATAC",
      ident.1 = null_levels[1],
      ident.2 = null_levels[2],
      test.use = "wilcox",
      min.pct = 0,
      logfc.threshold = 0,
      only.pos = FALSE
    )

    null_markers$peak <- rownames(null_markers)

    null_markers
  }
)

names(null_markers_atac_list) <- names(null_signac)

nullScores_atac_raw <- lapply(
  null_markers_atac_list,
  function(null_markers) {

    scores <- null_markers$p_val

    names(scores) <- rownames(null_markers)

    scores[
      is.finite(scores)
    ]
  }
)
```

After that, we use ClusterDE to call DEG results on the ATAC data.
```{r}
original_markers_atac$peak <- rownames(
  original_markers_atac
)

targetScores_atac <- original_markers_atac$p_val

names(targetScores_atac) <- rownames(
  original_markers_atac
)

de_atac <- ClusterDE::callDE(
  targetScores = targetScores_atac,
  nullScores = nullScores_atac_raw,
  FDR = 0.05,
  nlogTrans = TRUE,
  threshold = "DS"
)

print(
  paste0(
    "ClusterDE found ",
    sum(de_atac$record >= 0.5),
    " DA peaks"
  )
)
```

# Session information
```{r}
sessionInfo()
```
