---
title: "Perform ClusterDE on a PBMC dataset"
author: 
  - name: Dongyuan Song
    affiliation:
    - Bioinformatics IDP, University of California, Los Angeles
    email: dongyuansong@ucla.edu
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
package: "`r pkg_ver('ClusterDE')`"
vignette: >
  %\VignetteIndexEntry{ClusterDE-PBMC_vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE, warning=FALSE, results='hide'}
tools::R_user_dir("ClusterDE", which="cache")
library(ClusterDE)
library(Seurat)
library(SingleCellExperiment)
library(ggplot2)
library(patchwork)
library(SeuratData)
library(Matrix)
```

## Download data
The PBMC datasets are from `SeuratData`. We use one of them (10x Chromium (v3) from PBMC1 replicate). We filtered out some lowly epxressed genes to save computational time here.
```{r, cache=TRUE}
InstallData("pbmcsca")
data("pbmcsca")
pbmc <- pbmcsca[, pbmcsca@meta.data$Method=="10x Chromium (v3)" & pbmcsca@meta.data$Experiment == "pbmc1"]

pbmc <- pbmc[Matrix::rowSums(pbmc@assays$RNA@counts) > 100, ]
```

## Run the regular Seurat pipeline

We perform the default Seurat clustering. Note that in real data analysis, the cell type label is usually unknown.

```{r, fig.width=12, fig.height=6, cache=TRUE}
RNGkind("L'Ecuyer-CMRG")
set.seed(123)
pbmc <- NormalizeData(object = pbmc)
pbmc <- FindVariableFeatures(object = pbmc)
pbmc <- ScaleData(object = pbmc)
pbmc <- RunPCA(object = pbmc)
pbmc <- FindNeighbors(object = pbmc)
pbmc <- FindClusters(object = pbmc, resolution = 0.3)
pbmc <- RunUMAP(object = pbmc, dims = 1:10)
DimPlot(object = pbmc, reduction = "umap") + ggtitle("Clustering result") +
DimPlot(object = pbmc, reduction = "umap", group.by = "CellType")
```

In this vignette, we are interested in cluster 2 vs 8, which approximately represent CD14+/CD16+ monocytes. Please note that ClusterDE is designed for 1 vs 1 comparison. Therefore, users may (1) choose the two interested clusters manually based on their knowledge or (2) use the two closest clusters from computation (e.g., `BuildClusterTree` in `Seurat`).
```{r}
pbmc <- BuildClusterTree(pbmc)
PlotClusterTree(pbmc)
```

We perform the DE test between cluster 2 and 8. We subset the cluster 2 and 8 (`pbmc_sub`).
```{r}
pbmc_sub <- subset(x = pbmc, idents = c(2, 8))
original_markers <- FindMarkers(pbmc_sub, 
                                ident.1 = 2, 
                                ident.2 = 8, 
                                min.pct = 0, 
                                logfc.threshold = 0)
```

## Generate synthetic null data

We first generate the synthetic null data based on the target data (real subset data, `pbmc_sub`, which contains the two clusters you are interested in). You can increase the number of cores to speed it up.
```{r, cache=TRUE}
count_mat <- GetAssayData(object = pbmc_sub, slot = "counts")
set.seed(123)
synthetic_null <- ClusterDE::constructNull(count_mat, nCores = 2)
```

We perform the same pipeline as we did for target data. Please note we need two clusters here, too. This is straightforward for clustering method with predefined cluster numbers, e.g., K-means. For louvain clustering used in Seurat you may manually adjust the resolution parameter so that we still get two clusters.
```{r, cache=TRUE}
pbmc_null <- CreateSeuratObject(counts = synthetic_null)

set.seed(123)
pbmc_null <- NormalizeData(object = pbmc_null)
pbmc_null <- FindVariableFeatures(object = pbmc_null)
pbmc_null <- ScaleData(object = pbmc_null)
pbmc_null <- RunPCA(object = pbmc_null)
pbmc_null <- FindNeighbors(object = pbmc_null)
pbmc_null <- FindClusters(object = pbmc_null, resolution = 0.3)
pbmc_null <- RunUMAP(object = pbmc_null, dims = 1:10)
DimPlot(object = pbmc_null, reduction = "umap")
```

We perform the DE test on synthetic null data.
```{r}
null_markers <- FindMarkers(pbmc_null, 
                            ident.1 = 0, 
                            ident.2 = 1, 
                            min.pct = 0, 
                            logfc.threshold = 0)
```

We extract the p-values from both original data and synthetic null data, then use `ClusterDE` to "compare" them.
```{r}
original_pval <- original_markers$p_val
names(original_pval) <- rownames(original_markers)

null_pval <- null_markers$p_val
names(null_pval) <- rownames(null_markers)

res <- ClusterDE::callDE(original_pval, null_pval, nlogTrans = TRUE)
```

You can check the summary table. The genes are ranked by "contrast score (cs)".
```{r}
head(res$summaryTable)
```

To compare the result from the naive Seurat pipeline and ClusterDE, we first visualize the top 6 DE genes from Seurat. Genes *LST1* and *RPS19* are both highly expressed in two clusters. In addition, *RPS19* is reported as a stable housekeeping genes in several studies. 
```{r, fig.width=10, fig.height=6}
FeaturePlot(pbmc, features = c(rownames(original_markers)[1:6]), ncol = 3)
```
In contrast, the genes from ClusterDE do not have *LST1* and *RPS19* anymore.
```{r, fig.width=10, fig.height=6}
FeaturePlot(pbmc, features = c(res$summaryTable$Gene[1:6]), ncol = 3)
```

## Session information
```{r}
sessionInfo()
```
