---
title: "Perform ClusterDE on a microbiome dataset"
author: 
  - name: Yiwen Yang
    email: yyang5@fredhutch.org
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
package: "`r pkg_ver('ClusterDE')`"
vignette: >
  %\VignetteIndexEntry{ClusterDE-microbiome_vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
tools::R_user_dir("ClusterDE", which = "cache")
```

# Download data

Load lung cancer WGS microbiome profile processed in Seurat object, originally available at <https://github.com/yge15/TCGA_Microbial_Content>.
```{r}
data(lung, package = "ClusterDE")
```

# Generate synthetic null data

Generate 20 null replicates using ClusterDE stable to eliminate randomness of null generation. Note that for microbiome data pass the meta column name that stores log sequencing depth to both formula and other_covariates.
```{r}
RNGkind("L'Ecuyer-CMRG")
seed <- 123
null_data <- ClusterDE::constructNull(
  lung,
  data_type = "microbiome",
  formula = "log_depth",
  other_covariates = "log_depth",
  n_rep = 20,
  seed = seed
)
```

# Cell clustering

Perform Kmeans (K = 2) clustering on both real data and synthetic null data, the clustering method can be any method that is used in the original study.
```{r}
set.seed(seed)
clusters <- kmeans(t(Seurat::GetAssayData(lung, layer = "data")), 2)$cluster
null_clusters <- lapply(null_data, function(mat) {
  set.seed(seed)
  mat <- apply(mat, 2, function(x) x / sum(x))
  kmeans(t(mat), 2)$cluster
})
```

# DE analysis

Use any established DA methods to find differentially abundant taxa in both real and synthetic data. Using Wilcoxon, we can find 70 DA comparing the two clusters:
```{r}
original_pval <- apply(Seurat::GetAssayData(lung, layer = "data"), 1, function(x) {
  stats <- wilcox.test(x ~ clusters)
  stats$p.value
})
original_pval <- p.adjust(original_pval, method = "BH")
print(paste0("Wilcox found ", sum(original_pval < 0.05), " DA"))
```

Next, apply the same DA method on all null replicates.
```{r}
null_pvals <- lapply(seq_along(null_data), function(i) {
  null_pval <- apply(null_data[[i]], 1, function(x) {
    stats <- wilcox.test(x ~ null_clusters[[i]])
    stats$p.value
  })
  null_pval <- p.adjust(null_pval, method = "BH")
})
```

Finally, compare original p-values and null p-values using ClusterDE FDR control, we observed 12 real DA:
```{r}
da <- ClusterDE::callDE(original_pval, null_pvals)
print(paste0("ClusteDE found ", sum(da$record >= 0.5), " DA"))
```

# Session information
```{r}
sessionInfo()
```
