---
title: "Perform ClusterDE on a bulk microarray dataset"
author:
  - name: Pan Liu
    email: pliu3@fredhutch.org
  - name: Yiwen Yang
    email: yyang5@fredhutch.org
output:
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
package: "`r pkg_ver('ClusterDE')`"
vignette: >
  %\VignetteIndexEntry{ClusterDE-bulk-microarray_vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
tools::R_user_dir("ClusterDE", which = "cache")
```

# Download data

Load glioblastoma (GBM) dataset processed in Seurat object, originally from [Verhaak et al.](http://www.cell.com/cancer-cell/abstract/S1535-6108%2809%2900432-2) and can be downloaded at <https://gdc.cancer.gov/about-data/publications/gbm_exp>. Here we subset the Classical and Mesenchymal subtypes as an example.
```{r}
data(gbm_sub, package = "ClusterDE")
table(gbm_sub$subtype)
```

# Generate synthetic null data

Generate 20 null replicates using ClusterDE stable to eliminate randomness of null generation. Note that for normalized bulk microarray data, the distribution family should be "gaussian".
```{r, results = "hide", message = F}
RNGkind("L'Ecuyer-CMRG")
seed <- 123
null_data <- ClusterDE::constructNull(
  gbm_sub,
  data_type = "bulk_microarray",
  family = "gaussian",
  n_rep = 20,
  seed = seed
)
```

# Sample clustering

Use Consensus Clustering to divide the null samples into two clusters (following the same clustering procedure as the real data).
```{r, results = "hide", message = F, fig.keep = "none"}
set.seed(seed)
null_clusters <- lapply(null_data, function(mat) {
  mat <- as.matrix(mat)
  mat_centered <- mat - rowMeans(mat, na.rm = T)
  clusters <- ConsensusClusterPlus::ConsensusClusterPlus(
    mat_centered,
    maxK = 3,
    reps = 100,
    pItem = 0.8,
    pFeature = 1,
    clusterAlg = "hc",
    distance = "pearson",
    seed = seed,
    plot = F,
    title = tempdir()
  )[[2]]$consensusClass
})
```

# DE analysis

Use any established bulk methods to find DE genes in both real and null data. We use Wilcoxon as an example here.
```{r}
original_pval <- apply(Seurat::GetAssayData(gbm_sub, layer = "counts"), 1, function(x) {
  stats <- wilcox.test(x ~ gbm_sub$subtype)
  stats$p.value
})
original_pval <- p.adjust(original_pval, method = "BH")
print(paste0("Wilcox found ", sum(original_pval < 0.05), " DEG"))
```

```{r}
null_pvals <- lapply(seq_along(null_data), function(i) {
  null_pval <- apply(null_data[[i]], 1, function(x) {
    stats <- wilcox.test(x ~ null_clusters[[i]])
    stats$p.value
  })
  p.adjust(null_pval, method = "BH")
})
```

# FDR control

Finally, compare original p-values and null p-values using ClusterDE FDR control, we observed much fewer DEGs comparing to Wilcoxon:
```{r}
da <- ClusterDE::callDE(original_pval, null_pvals)
print(paste0("ClusteDE found ", sum(da$record >= 0.5), " DEG"))
```

# Session information

```{r}
sessionInfo()
```
