---
title: "Perform ClusterDE on a two-domain spatial dataset"
author: 
  - name: Siqi Chen
    affiliation:
    - Computer Science, Central South University
    email: siqichen4477@gmail.com
  - name: Dongyuan Song
    affiliation:
    - Bioinformatics IDP, University of California, Los Angeles
    email: dongyuansong@ucla.edu
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
package: "`r pkg_ver('ClusterDE')`"
vignette: >
  %\VignetteIndexEntry{ClusterDE-twodomains_vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
tools::R_user_dir("ClusterDE", which="cache")
```

```{r setup, message=FALSE, warning=FALSE, results='hide'}
#tools::R_user_dir("ClusterDE", which="cache")
library(Seurat)
library(SingleCellExperiment)
library(spatialLIBD)
library(ggplot2)
library(ClusterDE)
library(BayesSpace)
library(dplyr)
library(scales)
```

# Download data

We selected adjacent layer 6 and WM of 151673 slice from the LIBD Human Dorsolateral Prefrontal Cortex (DLPFC) dataset, which is downloaded in the spatialLIBD R package. We removed the genes expressed in fewer than 20% spots to save computational time here.

```{r}
### Download the spot-level data
spe <- fetch_data(type = "spe")

### Select the Layer6 and WM domains in the slice "151673"
sub_151673 <- spe[, spe$sample_id == "151673"]
index <- sub_151673$spatialLIBD == "L6"|sub_151673$spatialLIBD == "WM"
index[which(is.na(index))] <- "NAN"
sub_151673 <- sub_151673[, index=="TRUE"]
print(sub_151673)

# Delete the genes that express rate less than 20%
data <- sub_151673@assays@data$counts
zero_expre_rate <- apply(data,1,function(x){
  zero_true<-x==0
  zero_num<-length(which(zero_true==TRUE))/dim(data)[2]
  return(zero_num)
})
zero_expre_gene_idx <- which(zero_expre_rate<0.8)
sub_151673 <- sub_151673[zero_expre_gene_idx,]
cat(paste0("The size of data: ",dim(sub_151673)[1],"*",dim(sub_151673)[2],sep=""))

###construct the SingleCellExperiment object
real_sce <- SingleCellExperiment(list(counts=sub_151673@assays@data$counts))
###add colData information of singlecellexperiment
real_sce$spatial1 <- sub_151673@int_colData@listData$spatialCoords[,2]
real_sce$spatial2 <- sub_151673@int_colData@listData$spatialCoords[,1]
real_sce$cell_type <- sub_151673@colData$spatialLIBD
###visualize the real spatial domains
real_sce_domains <- data.frame(Xaxis=real_sce$spatial1,Yaxis=real_sce$spatial2,Domains=real_sce$cell_type)
ggplot(real_sce_domains, aes(x = Xaxis, y = Yaxis, col =Domains)) + geom_point(size=1.0) + coord_equal() +
  ggtitle("Manual annotation \n (The layer6 and WM in the slice 151673)") + 
  #scale_color_brewer(palette = "Accent") +
  theme(plot.title = element_text(size=10,hjust=0.5),
        panel.grid=element_blank(),
        panel.background = element_rect(fill = "gray90"),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 0.6),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank()) + scale_color_manual(values = c("#5791cc","#403f92"))
```

# Run the BayesSpace + Seurat pipeline

Firstly, we employed the BayesSpace for spatial clustering. Please note that ClusterDE is designed for 1 vs 1 comparison; therefore, we obtain two spatial clusters for illustration purpose.

```{r}
###construct the input of BayesSpace based on real dataset
#the input of BayesSpace is sce object
real_sce_clu <- SingleCellExperiment(list(counts=real_sce@assays@data$counts))
# add colData information of singlecellexperiment
real_sce_clu$row <- real_sce$spatial1
real_sce_clu$col <- real_sce$spatial2
## log-normalize the count data
set.seed(102)
real_sce_clu <- spatialPreprocess(real_sce_clu, platform="ST", n.PCs=7,log.normalize=TRUE)
#clustering with BayesSpace
set.seed(149)
real_sce_clu <- spatialCluster(real_sce_clu, q=2, platform="ST", d=7,
                               init.method="mclust", model="t", gamma=2,
                               nrep=1000, burn.in=100,
                               save.chain=TRUE)
#visualize the spatial cluster
real_sce_spatial_clu <- data.frame(Xaxis=real_sce_clu$row,Yaxis=real_sce_clu$col,Clusters=as.character(real_sce_clu$spatial.cluster))
ggplot(real_sce_spatial_clu, aes(x = Xaxis, y = Yaxis, col =Clusters)) + geom_point(size=1.0) + coord_equal() +
  ggtitle("Real data \n (spatial clusters detected by BayesSpace)") + 
  #scale_color_brewer(palette = "Accent") +
  theme(plot.title = element_text(size=10,hjust=0.5),
        panel.grid=element_blank(),
        panel.background = element_rect(fill = "gray90"),
        panel.border = element_rect(color = "black", fill = NA, size = 0.6),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank()) + scale_color_manual(values = c("#e87d72","#54bcc2"))
```

Then, we used the common DE  method (Wilcoxon Rank Sum Test) to identify domain marker genes between the two spatial clusters.

```{r}
#Identify domain marker genes in the real dataset based on the BayesSpace clustering result, follow Seurat tutorial
#create Seurat object
real_count_dataset <- real_sce@assays@data$counts
real_seurat <- CreateSeuratObject(counts = real_count_dataset, project = "real_seurat", min.cells = 0, min.features = 0)
real_seurat[["percent.mt"]] <- PercentageFeatureSet(real_seurat, pattern = "^MT-")
real_seurat <- NormalizeData(real_seurat, normalization.method = "LogNormalize", scale.factor = 10000)
real_seurat <- ScaleData(real_seurat, features = rownames(real_seurat))
real_ct <-real_sce_clu$spatial.cluster
names(real_ct) <- colnames(real_sce)
real_seurat[["cell_type"]] <- real_ct
Idents(real_seurat) <- "cell_type"
#Then we follow seurat tutorial to conduct DE analysis
real_markers <- FindMarkers(object = real_seurat, ident.1 = unique(real_ct)[1], ident.2 =  unique(real_ct)[2], test.use = "wilcox",logfc.threshold = 0, min.pct = 0, min.cells.feature = 1, min.cells.group = 1)
real_pvals <- real_markers[rownames(real_seurat),"p_val"]
names(real_pvals) <- rownames(real_seurat)
```

# Perform ClusterDE analysis

We first generate the synthetic null data based on the real data (target data). You can increase the number of cores to speed it up using the parameter "nCores".

```{r}
###generate synthetic null data
count_mat <- real_sce@assays@data$counts
location_mat <- data.frame(X=real_sce$spatial1,Y=real_sce$spatial2)
system.time(null_twodomain_dataset <- ClusterDE::constructNull(mat=count_mat,
                                                   family = "nb",
                                                   extraInfo=location_mat,
                                                   formula = "s(X, Y, bs = 'gp', k= 4)",
                                                   nCores = 2))
```

We perform the same pipeline as we did for target data. Please note we need two spatial clusters here, too. 

```{r}
###construct the input of BayesSpace based on null dataset######
null_sce <- SingleCellExperiment(list(counts =null_twodomain_dataset))
# add colData information of singlecellexperiment
null_sce$row <- real_sce$spatial1
null_sce$col <- real_sce$spatial2
## log-normalize the count data
set.seed(102)
null_sce<- spatialPreprocess(null_sce, platform="ST", n.PCs=7,log.normalize=TRUE)
#clustering with BayesSpace
set.seed(149)
null_sce <- spatialCluster(null_sce, q=2, platform="ST", d=7,
                           init.method="mclust", model="t", gamma=2,
                           nrep=1000, burn.in=100,
                           save.chain=TRUE)
#visualize the spatial cluster
null_spatial_clu <-  data.frame(Xaxis=null_sce$row,Yaxis=null_sce$col,Clusters=as.character(null_sce$spatial.cluster))
ggplot(null_spatial_clu, aes(x = Xaxis, y = Yaxis, col =Clusters))  + geom_point(size=1.0) + coord_equal() +
  ggtitle("Synthetic null data \n (spatial clusters detected by BayesSpace)") + 
  #scale_color_brewer(palette = "Accent") +
  theme(plot.title = element_text(size=10,hjust=0.5),
        panel.grid=element_blank(),
        panel.background = element_rect(fill = "gray90"),
        panel.border = element_rect(color = "black", fill = NA, size = 0.6),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank()) + scale_color_manual(values = c("#e87d72","#54bcc2"))
```
Visualize the spatial clustering results based on the synthetic null data.

Perform the common DE test (Wilcoxon Rank Sum Test) on synthetic null data.

```{r}
#Identify domain marker genes in the synthetic null dataset based on the BayesSpace clustering result, follow Seurat tutorial
#create Seurat object
null_count_dataset <- null_twodomain_dataset
null_seurat <- CreateSeuratObject(counts = null_count_dataset, project = "null_seurat", min.cells = 0, min.features = 0)
null_seurat[["percent.mt"]] <- PercentageFeatureSet(null_seurat, pattern = "^MT-")
null_seurat <- NormalizeData(null_seurat, normalization.method = "LogNormalize", scale.factor = 10000)
null_seurat <- ScaleData(null_seurat, features = rownames(null_seurat))
null_ct <- null_sce$spatial.cluster
names(null_ct) <- colnames(null_sce)
null_seurat[["cell_type"]] <- null_ct
Idents(null_seurat) <- "cell_type"
#Then we follow seurat tutorial to conduct DE analysis
null_markers <- FindMarkers(object = null_seurat, ident.1 = unique(null_ct)[1], ident.2 =  unique(null_ct)[2], test.use = "wilcox",logfc.threshold = 0, min.pct = 0, min.cells.feature = 1, min.cells.group = 1)
null_pvals <- null_markers[rownames(null_seurat),"p_val"]
names(null_pvals) <- rownames(null_seurat)
```

We extract the p-values from both target data and synthetic null data, then use `ClusterDE` to "compare" them.

```{r}
###you can set the value of FDR, the default value is 0.05
res <- ClusterDE::callDE(real_pvals, null_pvals, nlogTrans = TRUE)
cat(paste0("The number of domain marker genes is: ",length(res$DEgenes),sep=""))
```

We visualize the top DE genes from ClusterDE. As expected, the top genes detected by ClusterDE exhibit clear spatial expression patterns.

# Visualize top marker genes
```{r}
expression_dat_test <- data.frame(scales::rescale(as.matrix(t(log1p(counts(real_sce)))))) %>% as_tibble() %>% dplyr::mutate(X = colData(real_sce)$spatial1, Y = colData(real_sce)$spatial2) %>% tidyr::pivot_longer(-c("X", "Y"), names_to = "Gene", values_to = "Expression")

top_DE_ClusterDE <- res$DEgenes

p_topGenes <- expression_dat_test %>% as_tibble() %>% dplyr::mutate(Method = "ClusterDE") %>% filter(Gene %in% top_DE_ClusterDE[1:5]) %>% dplyr::mutate(Gene = factor(Gene, levels = top_DE_ClusterDE[1:5])) %>% ggplot(aes(x = X, y = Y, color = Expression)) + geom_point(size = 0.1) + scale_colour_gradientn(colors = viridis_pal(option = "magma")(10)) + coord_fixed(ratio = 1) + facet_grid(Method ~ Gene)+ theme_gray() + theme(axis.text.x = element_blank(), axis.text.y = element_blank(), axis.ticks.x = element_blank(), axis.ticks.y = element_blank(), legend.position = "bottom")

p_topGenes
```

# Session information

```{r}
sessionInfo()
```


